Analysis of cloud enhancement events in a 30-year record of global solar irradiance at Thessaloniki, Greece

### 2024-08-31 article/article.tex
Commit 2024-08-31 13:12

  \begin{abstract}
  In this study we investigate the characteristics of global horizontal
  irradiance enhancement events induced by clouds over Thessaloniki for
  the period 1993 -- 2023 using data recorded every one minute. We
  identified the cloud enhancement (CE) events by creating an
  appropriate clear-sky irradiance reference with the use of a radiative
  transfer model and aerosol optical depth data from a collocated Cimel
  sun photometer and a Brewer spectrophotometer. We found a trend in CE
  events of \(45.6\pm 21.9\,\text{cases}/\text{year}\), and a trend in
  the CE events irradiation of \(116.9\pm 67.8\,\text{kJ}/\text{year}\).
  The peak of the CE events was observed during May and June. The
  analysis of the total duration of CE events showed that durations
  longer than 5 minutes are very rare, with exceptions lasting over an
  hour and up to 140 minutes. Finally, we have detected enhancements
  above the total solar irradiance at the top of the atmosphere of up to
  \(400\,\text{W}/\text{m}^{2}\), with the \(75\,\%\) of the cases below
  \(200\,\text{W}/\text{m}^{2}\). The most active period of these
  extreme events is spring -- early summer with a secondary peak in
  autumn.
  \end{abstract}
    \begin{keyword}
    cloud enhancement \sep total solar radiation \sep global horizontal
irradiance \sep 
    over irradiance
  \end{keyword}

@@

We studied the enhancement of the global horizontal irradiance induced
by clouds over Thessaloniki for the period 1993 -- 2023 using data
recorded every one minute. We identified the cloud enhancement events
(CE) by creating an appropriate clear sky irradiance reference with the
use of a radiative transfer model and aerosol optical depth data from a
collocated Cimel sun photometer and a Brewer spectrophotometer. We found
a trend in CE events of \(46\pm 22\,\text{cases}/\text{year}\), and a
trend in the CE events irradiation of
\(116.3\pm 68.2\,\text{kJ}/\text{year}\). The peak of the CE events was
observed during May and June. An analysis of the total duration of CE
events, showed that durations longer than 5 minutes are very rare, with
exceptions lasting over an hour and up to 140 minutes. Finally, we have
detected enhancements above the total solar irradiance at the top of the
atmosphere for the same solar zenith angle of up to
\(400\,\text{W}/\text{m}^{2}\), with the \(75\,\%\) of the cases below
\(200\,\text{W}/\text{m}^{2}\). The most active season of these extreme
events is spring until early summer and autumn.

@@

To further improve the quality of the irradiance data, a manual
screening was performed, in order to remove inconsistent and erroneous
recordings that can occur stochastically or systematically during the
long operation of the instrument. The manual screening was aided by a
radiation data quality assurance procedure, adjusted for the site, which
was based on the methods of Long and Shi~\citep{Long2006, Long2008a}.
Thus, problematic recordings have been excluded from further processing.
Furthermore, due to the significant measurement uncertainty in GHI when
the Sun is near the horizon, and due to some systematic obstructions by
nearby buildings, we have excluded all measurements with solar zenith
angle (SZA) greater than \(78^\circ\). Finally, images from a sky camera
have been used in the manual inspection of the CE identification. The
sky camera operates since 2012 and stores images in 5 min time steps.

@@

The clear-sky reference values discussed above are based on the
climatological AOD and WC; hence they cannot describe accurately the
long-term variation of \(\text{GHI}_\text{ref}\) due to long-term
changes in the two atmospheric constituents, mainly AOD. As reported by
\citet{Natsis2023}, there is a long-term brightening effect in the GHI
data of Thessaloniki for the period 1993 -- 2023, which for clear-sky
data was attributed to long-term changes in aerosol effects. Therefore,
an adjustment of the \(\text{GHI}_\text{ref}\) during the period of
study was made using simulations with the RTM based on the long-term
variations of the AOD. As AERONET data start only in 2003, we used for
the period 1993 -- 2005 estimates of changes in AOD at
\(340\,\text{nm}\) derived from a collocated Brewer spectrophotometer
\citep{Kazadzis2007} to calculate the change in
\(\text{GHI}_\text{ref}\) due to aerosols during this period.

@@

\caption{Simulated long-term change in clear sky irradiance relative to the climatological values due to changes in AOD in Thessaloniki for the period 1993 -- 2023.}\label{fig:CS-change}
\end{figure}

@@

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the
yearly number of CE cases (each with duration of one minute), the yearly
mean OI and the yearly excess irradiation for the period 1993 -- 2023,
together with corresponding linear trends. All three quantities show
increasing trends, most pronounced for the frequency of occurrence
(\(+46\pm 22\,\text{cases}/\text{year}\)) and the excess irradiation
(\(+116.3 \pm 68.2\,\text{kJ}/\text{year}\)), which are also
statistically significant. In contrast the trend of the yearly mean OI
is negligible (\(+0.03\pm 0.1\,\text{W}/\text{m}^2/\text{year}\)) and of
no statistical significance. The average OI for the entire period is
\(+40.1\pm 2.5\,\text{W}/\text{m}^2\). The interannual variability of
the data about the trend lines is quite large. Furthermore, the spread
tends to increase with time (at least for the quantities of panels b and
c), suggesting a significant variability in cloud patterns over the
area, possibly associated to changes in climate.

@@

\caption{Seasonal variability of the number of CE events in Thessaloniki for the period 1993 -- 2023 normalized to the maximum occurring in June, in the form of a box and whisker plot. The monthly values have also been normalized to the relative abundance of valid GHI observations. The box contains the data between the lower $25\,\%$ and the upper $75\,\%$ percentiles, the thick horizontal line and the diamond symbol represent the median and the mean values, respectively. The vertical lines (whiskers) extend between the maximum and minimum monthly values.}\label{fig:relative-month-occurrences}
\end{figure}

@@

\caption{Relative frequency distribution of CE events in Thessaloniki for the period 1993 -- 2023 as a function of OI. The histogram was split in two plots with different y-axis scales for better readability.}\label{fig:ovir-distribution}
\end{figure}

@@

\caption{Relative frequency distribution of CE groups of consequent CE events according to their duration for Thessaloniki in the period 1993 -- 2023. The histogram was split in two plots with different y-axis scales for better readability.}\label{fig:ceg-duration-distribution}
\end{figure}

@@

\caption{Relation of excess irradiation of CE groups with their duration for Thessaloniki in the period 1993 -- 2023. The logarithmic color scale denotes the frequency of the respective groups of events.}\label{fig:group-2d}
\end{figure}

@@

\caption{Seasonal variability of the number of ECE events in Thessaloniki for the period 1993 -- 2023 normalized to the maximum occurring in March, in the form of a box and whisker plot. The box contains the data between the lower $25\,\%$ and the upper $75\,\%$ percentiles. The thick horizontal line and the diamond symbol represent the median and the mean values, respectively. The vertical lines (whiskers) extend between the maximum and minimum monthly values.}\label{fig:relative-month-occurancies-ECE}
\end{figure}

@@

\caption{Distribution of ECE events in Thessaloniki for the period 1993 -- 2023. Measured as extra irradiance over TSI at TOA for the same SZA.}\label{fig:P-extreme-distribution}
\end{figure}

@@

By creating a clear sky approximation representing the long- and
short-term variability of the expected clear sky GHI, we were able to
identify cases of CE events in Thessaloniki for the period 1993 -- 2023.
After analyzing the CE cases, we found an increase of
\(+45.6\pm 22\,\text{cases}/\text{year}\), with the mean annual
irradiation of the CE events increasing with a rate of
\(+116.3\pm 68.2\,\text{kJ}/\text{year}\). The most active months of CE
events are May and June. We found that continuous CE events can last up
to \(140\) minutes, while the duration of \(80\,\%\) of them is bellow
\(5\) minutes.

>>>>

  \begin{abstract}
  In this study we investigate the characteristics of global horizontal
  irradiance enhancement events induced by clouds over Thessaloniki for
  the period 1994 -- 2023 using data recorded every one minute. We
  identified the cloud enhancement (CE) events by creating an
  appropriate clear-sky irradiance reference with the use of a radiative
  transfer model and aerosol optical depth data from a collocated Cimel
  sun photometer and a Brewer spectrophotometer. We found a trend in CE
  events of \(45.6\pm 21.9\,\text{cases}/\text{year}\), and a trend in
  the CE events irradiation of \(116.9\pm 67.8\,\text{kJ}/\text{year}\).
  The peak of the CE events was observed during May and June. The
  analysis of the total duration of CE events showed that durations
  longer than 5 minutes are very rare, with exceptions lasting over an
  hour and up to 140 minutes. Finally, we have detected enhancements
  above the total solar irradiance at the top of the atmosphere of up to
  \(400\,\text{W}/\text{m}^{2}\), with the \(75\,\%\) of the cases below
  \(200\,\text{W}/\text{m}^{2}\). The most active period of these
  extreme events is spring -- early summer with a secondary peak in
  autumn.
  \end{abstract}
    \begin{keyword}
    cloud enhancement \sep total solar radiation \sep global horizontal
irradiance \sep 
    over irradiance
  \end{keyword}

@@

We studied the enhancement of the global horizontal irradiance induced
by clouds over Thessaloniki for the period 1994 -- 2023 using data
recorded every one minute. We identified the cloud enhancement events
(CE) by creating an appropriate clear sky irradiance reference with the
use of a radiative transfer model and aerosol optical depth data from a
collocated Cimel sun photometer and a Brewer spectrophotometer. We found
a trend in CE events of \(+46\pm 22\,\text{cases}/\text{year}\), and a
trend in the CE events irradiation of
\(+116.3\pm 68.2\,\text{kJ}/\text{year}\). The peak of the CE events was
observed during May and June. An analysis of the total duration of CE
events, showed that durations longer than 5 minutes are very rare, with
exceptions lasting over an hour and up to 140 minutes. Finally, we have
detected enhancements above the total solar irradiance at the top of the
atmosphere for the same solar zenith angle of up to
\(400\,\text{W}/\text{m}^{2}\), with the \(75\,\%\) of the cases below
\(200\,\text{W}/\text{m}^{2}\). The most active season of these extreme
events is spring until early summer and autumn.

@@

To further improve the quality of the irradiance data, a manual
screening was performed, in order to remove inconsistent and erroneous
recordings that can occur stochastically or systematically during the
long operation of the instrument. The manual screening was aided by a
radiation data quality assurance procedure, adjusted for the site, which
was based on the methods of Long and Shi~\citep{Long2006, Long2008a}.
Thus, problematic recordings have been excluded from further processing.
Furthermore, due to the significant measurement uncertainty in GHI when
the Sun is near the horizon, and due to some systematic obstructions by
nearby buildings, we have excluded all measurements with solar zenith
angle (SZA) greater than \(78^\circ\). Finally, images from a sky camera
have been used in the manual inspection of the CE identification. The
sky camera operates since 2012 and stores images in 5 min time steps. An
overview of the data analyzed for the study of the CE and ECE events is
given in Figure\nobreakspace{}\ref{fig:CLB-daily}, as the daily mean of
the measured GHI.

{\centering \includegraphics[width=1\linewidth]{../images/P_daily_trend-3}

\caption{Daily mean values of GHI in Thessaloniki for the period 1994 -- 2023.}\label{fig:CLB-daily}
\end{figure}

@@

The clear-sky reference values discussed above are based on the
climatological AOD and WC; hence they cannot describe accurately the
long-term variation of \(\text{GHI}_\text{ref}\) due to long-term
changes in the two atmospheric constituents, mainly AOD. As reported by
\citet{Natsis2023}, there is a long-term brightening effect in the GHI
data of Thessaloniki for the period 1993 -- 2023, which for clear-sky
data was attributed to long-term changes in aerosol effects. Therefore,
an adjustment of the \(\text{GHI}_\text{ref}\) during the period of
study was made using simulations with the RTM based on the long-term
variations of the AOD. As AERONET data start only in 2003, we used for
the period 1994 -- 2005 estimates of changes in AOD at
\(340\,\text{nm}\) derived from a collocated Brewer spectrophotometer
\citep{Kazadzis2007} to calculate the change in
\(\text{GHI}_\text{ref}\) due to aerosols during this period.

@@

\caption{Simulated long-term change in clear sky irradiance relative to the climatological values due to changes in AOD in Thessaloniki for the period 1994 -- 2023.}\label{fig:CS-change}
\end{figure}

@@

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the
yearly number of CE cases (each with duration of one minute), the yearly
mean OI and the yearly excess irradiation for the period 1994 -- 2023,
together with corresponding linear trends. All three quantities show
increasing trends, most pronounced for the frequency of occurrence
(\(+46\pm 22\,\text{cases}/\text{year}\)) and the excess irradiation
(\(+116.3 \pm 68.2\,\text{kJ}/\text{year}\)), which are also
statistically significant. In contrast the trend of the yearly mean OI
is negligible (\(+0.03\pm 0.1\,\text{W}/\text{m}^2/\text{year}\)) and of
no statistical significance. The average OI for the entire period is
\(+40.1\pm 2.5\,\text{W}/\text{m}^2\). The interannual variability of
the data about the trend lines is quite large. Furthermore, the spread
tends to increase with time (at least for the quantities of panels b and
c), suggesting a significant variability in cloud patterns over the
area, possibly associated to changes in climate.

@@

\caption{Seasonal variability of the number of CE events in Thessaloniki for the period 1994 -- 2023 normalized to the maximum occurring in June, in the form of a box and whisker plot. The monthly values have also been normalized to the relative abundance of valid GHI observations. The box contains the data between the lower $25\,\%$ and the upper $75\,\%$ percentiles, the thick horizontal line and the diamond symbol represent the median and the mean values, respectively. The vertical lines (whiskers) extend between the maximum and minimum monthly values.}\label{fig:relative-month-occurrences}
\end{figure}

@@

\caption{Relative frequency distribution of CE events in Thessaloniki for the period 1994 -- 2023 as a function of OI. The histogram was split in two plots with different y-axis scales for better readability.}\label{fig:ovir-distribution}
\end{figure}

@@

\caption{Relative frequency distribution of CE groups of consequent CE events according to their duration for Thessaloniki in the period 1994 -- 2023. The histogram was split in two plots with different y-axis scales for better readability.}\label{fig:ceg-duration-distribution}
\end{figure}

@@

\caption{Relation of excess irradiation of CE groups with their duration for Thessaloniki in the period 1994 -- 2023. The logarithmic color scale denotes the frequency of the respective groups of events.}\label{fig:group-2d}
\end{figure}

@@

\caption{Seasonal variability of the number of ECE events in Thessaloniki for the period 1994 -- 2023 normalized to the maximum occurring in March, in the form of a box and whisker plot. The box contains the data between the lower $25\,\%$ and the upper $75\,\%$ percentiles. The thick horizontal line and the diamond symbol represent the median and the mean values, respectively. The vertical lines (whiskers) extend between the maximum and minimum monthly values.}\label{fig:relative-month-occurancies-ECE}
\end{figure}

@@

\caption{Distribution of ECE events in Thessaloniki for the period 1994 -- 2023. Measured as extra irradiance over TSI at TOA for the same SZA.}\label{fig:P-extreme-distribution}
\end{figure}

@@

By creating a clear sky approximation representing the long- and
short-term variability of the expected clear sky GHI, we were able to
identify cases of CE events in Thessaloniki for the period 1994 -- 2023.
After analyzing the CE cases, we found an increase of
\(+45.6\pm 22\,\text{cases}/\text{year}\), with the mean annual
irradiation of the CE events increasing with a rate of
\(+116.3\pm 68.2\,\text{kJ}/\text{year}\). The most active months of CE
events are May and June. We found that continuous CE events can last up
to \(140\) minutes, while the duration of \(80\,\%\) of them is bellow
\(5\) minutes.
